% A ConTeXt document [master document: commdiag.tex]

\startchapter[title=Overview]

This document defines the Commutative Diagrams for \ConTeXt\ module, 
\type{t-commdiag}. This module provides (Mathematical) Commutative 
Diagrams using \ConTeXt's underlying MetaPost/MetaFun graphics primitives. 

This document uses the \type{t-literate-modules} \ConTeXt\ module to 
define the \type{t-commdiag} module. This means that the code required to 
implement the Commutative Diagrams module is an integral part of this document. 

As far as the \type{t-commdiag} module is concerned, a commutative diagram 
is a collection of \quote{objects} laid out in a (rectangular) martix, 
together with a collection of \quote{arrows} between pairs of 
\quote{objects}. The label for a given object is typeset using \ConTeXt\ 
and is then drawn in the appropriate location in the matrix. Each arrow 
can be drawn with a number of arrow bodies, as well as arrow heads and/or 
tails. Each arrow can also have a label, again typeset in \ConTeXt. 

\section[title=Using \type{t-commdiag}]

To \emph{use} the \type{t-commdiag} module, you need to \type{use} the 
\type{t-commdiag} module in the \quote{usual} \ConTeXt\ way: 

\starttyping
\usemodule[t-commdiag]
\stoptyping

\noindent\ This \type{\usemodule} statement must occur in the document's 
setup section.

\startbuffer
\startformula \startMPcode{commDiag}  
  setupCommDiags ; 
  
  addObject(1,1, "X_0");
  addObject(1,2, "X_1");
  
  addObject(2,1, "Y_0");
  addObject(2,2, "Y_1");

  drawObjects(1.75cm, 2cm);

  addArrow(1,1, 1,2, ">", 0)()()("j_0",     0.5, top);
  addArrow(1,1, 2,1, ">", 0)()()("k_0",     0.5, lft);
  addArrow(1,2, 2,2, ">", 0)()()("k_1",     0.5, rt);
  addArrow(2,1, 2,2, ">", 0)()()("t_1",     0.5, bot);
\stopMPcode \stopformula
\stopbuffer

As a simple example, consider the following square commutative diagram: 
\processTEXbuffer

\noindent\ It was created using the following \type{t-commdiag} code:

\typebuffer

\section[title=Describing a commutative diagram]

A typical \type{t-commdiag} diagram's code is enclosed in a \ConTeXt\ 
MetaFun environment (in this case the pair of \type{\startMPcode} and 
\type{\stopMPcode} \ConTeXt\ macros). While a \type{t-commdiag} 
commutative diagram \emph{can} be position inline (inside running text), 
it is typically positioned using one of \ConTeXt's displayed mathematics 
or figure environments (in this case the pair of \type{\startformula} and 
\type{\stopformula} \ConTeXt\ macros). 

The \type{t-commdiag} code which describes a given commutative diagram has 
five sections: 

\startitemize[n]

\item {\bf The \type{setupComDiags} statement.} The \type{setupCommDiags} 
statement sets up the \type{objLabel} (object labels), and \type{objPos} 
(object position) matrices, as well as the \type{numRows} and 
\type{numCols} variables. These \type{objLabel} and \type{objPos} matrices 
are used by the \type{addObject} and \type{addArrow} statements, to 
correctly align the objects and arrows of the commutative diagram. These 
matrices can also be used in your own MetaPost/MetaFun code. 

\item {\bf A collection of \type{addObject} statements.} The 
\type{addObject} macro takes three arguments. The first two are the row 
and column in the overall commutative diagram's matrix in which the object 
is located. The third and last argument is the object label as a string, 
which is typeset by \ConTeXt. The row and column are expected to be 
integers and are indices into the \type{objLabel} and \type{objPos} 
matrices. You are not required to supply an object for every row, column 
combination. Matrix cells which have no object label are simply not 
drawn\footnote{The object position values \emph{are} computed should they 
be needed in your own code.}. 

\item {\bf The \type{drawObjects} statement.} The two arguments to the 
\type{drawObjects} macro, are the row height, and column width 
respectively. You can use these two parameters to adjust the overall size 
of your diagram. The \type{drawObjects} statement sets the locations of 
all of the object labels and then draws them\footnote{MetaPost has the 
native ability to compute layouts given a collection of drawing 
constraints. However, at this time, we keep things simple by 
\emph{explicitly} supplying the row height and column width. At sometime 
in the future, we may implement full \emph{implicit} layout using 
MetaPost's native constraint solving capabilities.}. You can use these 
object labels or object positions in your own MetaPost/MetaFun code. 

\item {\bf A collection of \type{addArrow} statements.} The 
\type{addArrow} macro is rather more complex and is discussed below. 

\item {\bf Optionally, your own MetaPost/MetaFun code.} If for some reason 
the \type{t-commdiag} primitives are not sufficient for your needs, you 
are able to use the \type{objLabel} and \type{objPos} matrices in your own 
MetaPost/MetaFun code. However your own code \emph{must} be located 
\emph{after} the \type{drawObjects} statement, and preferably after all 
\type{addArrow} statements. 

\stopitemize

\noindent\ The order of these four sections is \emph{critical}, they will 
not work in any other order.

\section[title=The \type{addArrow} macro]

{\bf The \type{addArrow} macro} has upto 12 arguments in a number of 
logical groups. The {\bf arrow is described} by the first six arguments. 
The next two arguments, as well as the last (optional) argument 12, all 
provide you the ability to specify {\bf MetaPost/MetaFun picture 
commands}. Arguments 9, 10 and 11 describe the arrow's label. 

\startitemize[n]

\item {\bf Arrow description } \type{fromRow} integer index. 

\item {\bf Arrow description } \type{fromCol} integer index. 

\item {\bf Arrow description } \type{toRow} integer index. 

\item {\bf Arrow description } \type{toCol} integer index. Together these 
four indices, \type{(fromRow, fromCol)} and \type{(toRow, toCol)}, specify 
the objects which the arrow connects. 

\item {\bf Arrow description } \type{arrowDecorations} is a string which 
describes the tail, body and head of the arrow. Its contents and meaning 
are described below. 

\item {\bf Arrow description } \type{bend} a numerical angle to the left 
of the straight line with which the line should bend away from the 
straight line. Negative angles bend to the right. 

\item {\bf MetaPost/MetaFun picture command} (optional) arrow body 
modifiers. These MetaPost/MetaFun command \emph{must} be surrounded by a 
pair of round brackets, \type{"("} and \type{")"}. If you only supply an 
empty pair of round brackets, no commands will be used. 

\item {\bf MetaPost/MetaFun picture command} (optional) arrow tail and/or 
head modifiers. These MetaPost/MetaFun command \emph{must} be surrounded 
by a pair of round brackets, \type{"("} and \type{")"}. If you only supply 
an empty pair of round brackets, no commands will be used. 

\item {\bf Label description} a label string which is typeset by \ConTeXt. 
If the label string is empty, \type{""}, no label will be drawn.

\item {\bf Label description} the linear label position (a numeric value 
between zero and one), which denotes how far along the arrow to place the 
label. Values less than or equal to zero or greater than or equal to one 
will place the label in the \emph{middle} of the arrow. 

\item {\bf Label description} the relative label position (a symbolic 
value) indicates where the label should be placed relative to the position 
along the arrow's length. You can use any value, such as \type{lft}, 
\type{rt}, \type{top}, or \type{bot}, which can be used with the MetaPost 
\type{label} macro. 

\item {\bf MetaPost/MetaFun picture command} (optional) modifies the 
\emph{whole} arrow. In typical MetaPost/MetaFun fashion, these modifiers 
are \emph{outside} of the \type{addArrow}'s formal arguments (that is they 
are everything between the last \type{")"} and the statement's \type{";"}). 

\stopitemize


{\bf The \type{arrowDecorations} string} consists of three parts. The 
first (optional) part describes the tail decorations. The middle 
(optional) part describes the type of arrow. The last (optional) part 
describes the head decorations. An empty, \type{""}, 
\type{arrowDecorations} string will be replaced by the string \type{"->"}. 
The tail and head decorations are deliniated by the \emph{first} body 
decoration\footnote{Any body decorators other than the first one found 
will be ignored.}, \type{"-"}, \type{"="}, \type{"~"}, or \type{"^"}, 
found in the string. If no body decoration is found, the tail decorations 
are assumed to be \type{""} and the body decorations are \type{"-"}. In 
this case all of the \type{arrowDecroations} are used to describe the 
arrow heads. The tail and head decorations are: 

\startitemize

\item \type{"o"} a dot at the tail or head end of the arrow.

\item \type{">"} an arrow head at the tail or head end of the arrow, 
directed in the positive direction of the arrow. 

\item \type{"<"} an arrow head at the tail or head end of the arrow, 
directed in the opposite direction of the arrow.

\item \type{"]"} a second arrow head at the tail or head 
end of the arrow, directed in the positive direction of the arrow. 

\item \type{"["} a second arrow head at the tail or head 
end of the arrow, directed in the opposite direction of the arrow. 

\stopitemize

\noindent\ The body decorations are:

\startitemize

\item \type{"-"} the arrow will be drawn with a simple straight line.

\item \type{"="} the arrow will be drawn with a double \quote{equals} 
line. 

\item \type{"~"} the arrow will be drawn with a squiggly line.

\item \type{"^"} the arrow will be drawn with a zig zag line.

\stopitemize

\section[title=Examples]

\useURL[guideCDPackages][https://www.jmilne.org/not/CDGuide.html]

The best way to understand how these macros can be used is to read through 
the following examples. These examples are taken from James Milne's 
\quote{Guide to Commutative Diagram Packages} which can be found at 
\from[guideCDPackages]. In particular Milne's guide to drawing commutative 
diagrams using the TikZ and/or TikZ-cd packages has the greatest number of 
complex diagrams. We will provide examples of how to use \type{t-commdiag} 
to draw each of these examples.

\startbuffer
\startformula\startMPcode{commDiag}
  setupCommDiags ;
  
  addObject(1,1, "A") ;
  addObject(1,2, "B") ;
  addObject(1,3, "C") ;
  addObject(2,2, "D") ;
  
  drawObjects(1.75cm, 2cm);
  
  addArrow(1,1, 1,2, ">->", 0)()()("u", 0.5, top);
  addArrow(1,1, 1,2, ">->", 0)()()("b", 0.5, bot);

  addArrow(1,3, 1,2, ">->", 0)()()("u", 0.5, top);
  addArrow(1,3, 1,2, ">->", 0)()()("b", 0.5, bot);
  
  addArrow(1,1, 2,2, "]>",  0)()()("u", 0.5, urt);
  addArrow(1,1, 2,2, "]>",  0)()()("b", 0.5, llft);
  
  addArrow(1,3, 2,2, "]>",  0)()()("u", 0.5, ulft);
  addArrow(1,3, 2,2, "]>",  0)()()("b", 0.5, lrt);

  addArrow(1,2, 2,2, ">",   0)
    (dashed withdots scaled 0.5)()("l", 0.5, lft);
  addArrow(1,2, 2,2, ">",   0)
    (dashed withdots scaled 0.5)()("r", 0.5, rt);
\stopMPcode\stopformula
\stopbuffer

\processTEXbuffer

\typebuffer

\startbuffer
\startformula\startMPcode{commDiag}
  setupCommDiags ;
  
  addObject(1,1, "A") ;
  addObject(1,2, "B") ;
  addObject(1,3, "C") ;
  addObject(2,1, "D") ;
  addObject(2,2, "E") ;
  addObject(2,3, "F") ;
  
  drawObjects(1.75cm, 4cm);
  
  addArrow(1,1, 1,2, ">", 0)()()("", 0.5, );
  addArrow(1,2, 1,3, ">", 0)()()("\text{very long label}", 0.5, top);
  
  addArrow(2,1, 2,2, ">", 0)()()("", 0.5, );
  addArrow(2,2, 2,3, ">", 0)()()("", 0.5, );
  
  addArrow(1,1, 2,1, ">", 0)()()("", 0.5, );
  addArrow(1,2, 2,2, ">", 0)()()("", 0.5, );
  addArrow(1,3, 2,3, ">", 0)()()("", 0.5, );
\stopMPcode\stopformula
\stopbuffer

\processTEXbuffer

\typebuffer

\startbuffer
\startformula\startMPcode{commDiag}
  setupCommDiags ;
  
  addObject(1,1, "\hat{A}") ;
  addObject(1,2, "\underset{n \in Z}{\prod} A_n") shifted (0, -5bp);
  addObject(1,3, "\underset{n \in Z}{\prod} A_n") shifted (0, -5bp);
  
  drawObjects(1.75cm, 2cm);
  
  addArrow(1,1, 1,2, ">", 0)()()("", 0.5, ) ;
  addArrow(1,2, 1,3, ">", 0)()()("", 0.5, ) ;
\stopMPcode\stopformula
\stopbuffer

\processTEXbuffer

\typebuffer

\startbuffer
\startformula\startMPcode{commDiag}
  setupCommDiags ;
  
  addObject(1,1, "A") ;
  addObject(1,2, "B") ;
  
  drawObjects(1.75cm, 2cm);
  
  addArrow(1,1, 1,2, ">",  60)()()("", 0.5, );
  addArrow(1,1, 1,2, ">", -60)()()("", 0.5, );
  
  save upperArrow, lowerArrow, anArrow;
  path upperArrow, lowerArrow, anArrow;
  %
  upperArrow := computeArrow(1,1, 1,2,  50);
  lowerArrow := computeArrow(1,1, 1,2, -50);
  anArrow    := 
    point 0.5*length(upperArrow) of upperArrow --
    point 0.5*length(lowerArrow) of lowerArrow;
  %
  drawFilledArrows := 0;
  drawCDArrow("=>", anArrow)()();
  drawFilledArrows := 1;
\stopMPcode\stopformula
\stopbuffer

\processTEXbuffer

\typebuffer

\startbuffer
\startformula\startMPcode{commDiag}
  setupCommDiags ;
  
  addObject(1,2, "A") ;
  
  addObject(2,2, "B") ;
  addObject(2,3, "C") ;
  addObject(2,4, "D") ;
  addObject(2,5, "E") ;
  
  addObject(3,1, "F") ;
  
  drawObjects(1.75cm, 2cm);
  
  addArrow(1,2, 2,3, ">", 0)()()("c", 0.5, );
  addArrow(1,2, 2,4, ">", 0)()()("d", 0.5, );
  addArrow(1,2, 2,5, ">", 0)()()("e", 0.5, );

  addArrow(2,2, 2,3, ">", 0)()()("", 0.5, );
  addArrow(2,3, 2,4, ">", 0)()()("", 0.5, );
  addArrow(2,4, 2,5, ">", 0)()()("", 0.5, );
  
  addArrow(1,2, 3,1, ">", 0)()()("f", 0.5, ulft);
  addArrow(2,2, 3,1, ">", 0)()()("",  0.5, );
\stopMPcode\stopformula
\stopbuffer

\processTEXbuffer

\typebuffer

\startbuffer
\startformula\startMPcode{commDiag}
  setupCommDiags ;
  
  addObject(1,1, "T") ;
  
  addObject(2,2, "X \times_Z Y") ;
  addObject(2,3, "X") ;
  
  addObject(3,2, "Y") ;
  addObject(3,3, "Z") ;
    
  drawObjects(1.75cm, 2.5cm);
  
  addArrow(1,1, 2,3, ">", 30)()()("x", 0.5, urt);
  addArrow(1,1, 2,2, ">", 0)
    (dashed withdots scaled 0.5)()("(x,y)", 0.5, );
  addArrow(1,1, 3,2, ">", -30)()()("y", 0.5, llft);

  addArrow(2,2, 2,3, ">", 0)()()("p", 0.5, top);
  addArrow(2,2, 3,2, ">", 0)()()("q", 0.5, rt);
  addArrow(2,3, 3,3, ">", 0)()()("f", 0.5, rt);
  addArrow(3,2, 3,3, ">", 0)()()("g", 0.5, top);
\stopMPcode\stopformula
\stopbuffer

\processTEXbuffer

\typebuffer

\startbuffer
\startformula\startMPcode{commDiag}
  setupCommDiags ;
  
  addObject(2,1, "0") ;
  addObject(2,2, "\text{Br}(L/K)") ;
  addObject(2,3, "\bigoplus_v \text{Br}(L^v/K_v)") ;
  
  addObject(1,4, "H^2(L/K)") ;
  addObject(3,4, "\mathbb{Q}/\mathbb{Z}") ;
    
  drawObjects(1.75cm, 3cm);
  
  addArrow(2,1, 2,2, ">", 0)()()("", 0.5, );
  addArrow(2,2, 2,3, ">", 0)()()("", 0.5, );
  
  addArrow(2,3, 1,4, ">", 0)()()("", 0.5, );
  addArrow(2,3, 3,4, ">", 0)()()("", 0.5, );
\stopMPcode\stopformula
\stopbuffer

\processTEXbuffer

\typebuffer

\startbuffer
\startformula\startMPcode{commDiag}
  setupCommDiags ;
  
  addObject(1,3, "X \otimes (Y \otimes (Z \otimes T))") ;
  
  addObject(2,1, "X \otimes ((Y \otimes Z) \otimes T)") ;
  addObject(2,5, "(X \otimes Y) \otimes (Z \otimes T)") ;
  
  addObject(3,2, "(X \otimes (Y \otimes Z)) \otimes T") ;
  addObject(3,4, "((X \otimes Y) \otimes Z) \otimes T") ;
    
  drawObjects(1.75cm, 2.5cm);
  
  addArrow(1,3, 2,1, ">", 0)()()("1 \otimes \phi", 0.5, ulft);
  addArrow(1,3, 2,5, ">", 0)()()("\phi", 0.5, urt);
  
  addArrow(2,1, 3,2, ">", 0)()()("\phi", 0.5, llft);
  addArrow(2,5, 3,4, ">", 0)()()("\phi", 0.5, lrt);
  
  addArrow(3,2, 3,4, ">", 0)()()("\phi \otimes 1", 0.5, top);
\stopMPcode\stopformula
\stopbuffer

\processTEXbuffer

\typebuffer

\startbuffer
\startformula\startMPcode{commDiag}
  setupCommDiags ;
  
  addObject(1,1, "Y \times_X Y") ;
  addObject(1,2, "Y") ;
  addObject(1,3, "X") ;
    
  drawObjects(1.75cm, 2.5cm);
  
  addArrow(1,1, 1,2, ">", 0)()()("p_1", 0.5, top) shifted (0, 3bp);
  addArrow(1,1, 1,2, ">", 0)()()("p_2", 0.5, bot) shifted (0, -3bp);
  
  addArrow(1,2, 1,3, ">", 0)()()("", 0.5, );
\stopMPcode\stopformula
\stopbuffer

\processTEXbuffer

\typebuffer

\startbuffer
\startformula\startMPcode{commDiag}
  setupCommDiags ;
  
  addObject(1,1, "\text{\sans S}(Z)") ;
  addObject(1,2, "\text{\sans S}(X)") ;
  addObject(1,3, "\text{\sans S}(U)") ;
    
  drawObjects(1.75cm, 2.5cm);
  
  addArrow(1,2, 1,1, ">", 0)()()("i^*", 0.5, top) shifted (0, 10bp);
  addArrow(1,1, 1,2, ">", 0)()()("i_*", 0.5,    ) ;
  addArrow(1,2, 1,1, ">", 0)()()("i^!", 0.5, bot) shifted (0, -10bp);
  
  addArrow(1,3, 1,2, ">", 0)()()("i^*", 0.5, top) shifted (0, 10bp);
  addArrow(1,2, 1,3, ">", 0)()()("i_*", 0.5,    ) ;
  addArrow(1,3, 1,2, ">", 0)()()("i^!", 0.5, bot) shifted (0, -10bp);
\stopMPcode\stopformula
\stopbuffer

\processTEXbuffer

\typebuffer

\startbuffer
\startformula\startMPcode{commDiag}
  setupCommDiags ;
  
  addObject(1,1, "0") ;
  addObject(1,2, "\text{Ker}\; f") ;
  addObject(1,3, "\text{Ker}\: a") ;
  addObject(1,4, "\text{Ker}\: b") ;
  addObject(1,5, "\text{Ker}\: b") ;

  addObject(2,3, "A") withcolor blue;
  addObject(2,4, "B") withcolor blue;
  addObject(2,5, "C") withcolor blue;
  addObject(2,6, "0") withcolor blue;
  
  addObject(3,2, "0")  withcolor blue;
  addObject(3,3, "A'") withcolor blue;
  addObject(3,4, "B'") withcolor blue;
  addObject(3,5, "C'") withcolor blue;

  addObject(4,3, "\text{Coker}\: a") ;
  addObject(4,4, "\text{Coker}\: b") ;
  addObject(4,5, "\text{Coker}\: c") ;
  addObject(4,6, "\text{Coker}\: g'") ;
  addObject(4,7, "0") ;
    
  drawObjects(1.75cm, 2cm);
  
  addArrow(1,1, 1,2, ">", 0)()()("", 0.5, ) withcolor red;
  addArrow(1,2, 1,3, ">", 0)()()("", 0.5, ) withcolor red;
  addArrow(1,3, 1,4, ">", 0)()()("", 0.5, ) withcolor red;
  addArrow(1,4, 1,5, ">", 0)()()("", 0.5, ) withcolor red;
  addArrow(1,5, 4,3, ">", 135)()()("d", 0.5, ulft) withcolor red;
  
  addArrow(1,3, 2,3, ">", 0)()()("", 0.5, );
  addArrow(1,4, 2,4, ">", 0)()()("", 0.5, );
  addArrow(1,5, 2,5, ">", 0)()()("", 0.5, );
  
  addArrow(2,3, 2,4, ">", 0)()()("f", 0.5, top) withcolor blue;
  addArrow(2,4, 2,5, ">", 0)()()("g", 0.5, top) withcolor blue;
  addArrow(2,5, 2,6, ">", 0)()()("", 0.5, ) withcolor blue;

  addArrow(2,3, 3,3, ">", 0)()()("a", 0.5, rt) withcolor blue;
  addArrow(2,4, 3,4, ">", 0)()()("b", 0.5, rt) withcolor blue;
  addArrow(2,5, 3,5, ">", 0)()()("c", 0.5, rt) withcolor blue;

  addArrow(3,2, 3,3, ">", 0)()()("", 0.5, ) withcolor blue;
  addArrow(3,3, 3,4, ">", 0)()()("f'", 0.5, bot) withcolor blue;
  addArrow(3,4, 3,5, ">", 0)()()("g'", 0.5, bot) withcolor blue;

  addArrow(3,3, 4,3, ">", 0)()()("", 0.5, );
  addArrow(3,4, 4,4, ">", 0)()()("", 0.5, );
  addArrow(3,5, 4,5, ">", 0)()()("", 0.5, );
  
  addArrow(4,3, 4,4, ">", 0)()()("", 0.5, ) withcolor red;
  addArrow(4,4, 4,5, ">", 0)()()("", 0.5, ) withcolor red;
  addArrow(4,5, 4,6, ">", 0)()()("", 0.5, ) withcolor red;
  addArrow(4,6, 4,7, ">", 0)()()("", 0.5, ) withcolor red;

\stopMPcode\stopformula
\stopbuffer

\processTEXbuffer

\typebuffer


\stopchapter 